---
title: "Ex. 04-2: Downstream ATAC-seq analysis"
output: 
    rmdformats::readthedown: 
        highlight: tango
        preserve_yaml: true
        df_print: tibble
        toc_depth: 4
        css: ../../../custom.css
    rmdformats::downcute: 
        preserve_yaml: true
    rmarkdown::html_document: 
        theme: simplex
        highlight: tango
        preserve_yaml: true
        code_folding: hide
        df_print: tibble
        toc: true
        toc_float: true
---

## 1. Import ATAC peaks from YAPC

YAPC has mapped peaks with several different thresholds of significance. 

> Check how many peaks were mapped at each threshold.  

<details><summary style="color: #ff7f00; font-weight: bold">Show code</summary><p>
```{sh}
wc -l Share/day04/_yapc-files/stress-response_*.bed
```
</p></details><br>

> Which type of genomic locus does an ATAC peak corresponds to? A priori, which YAPC significance threshold seems to yield the most appropriate number of peaks?  

<details><summary style="color: #ff7f00; font-weight: bold">Show code</summary><p>
```{r}
# Import ATAC peaks as GRanges
library(tidyverse)
ATAC_peaks <- read.table('Share/day04/_yapc-files/stress-response_0.05.bed') %>% 
    as_tibble() %>% 
    setNames(c('seqnames', 'start', 'end', 'yapc_scores', 'score', 'strand', 'summit', 'summit2')) %>%
    GenomicRanges::makeGRangesFromDataFrame(keep.extra.columns = TRUE)
ATAC_peaks$peakID <- paste0('peak_', 1:length(ATAC_peaks))
ATAC_peaks$yapc_scores <- NULL
ATAC_peaks$summit2 <- NULL
ATAC_peaks$score <- NULL
ATAC_peaks
```
</p></details><br>

## 2. Brief QC of ATAC peaks 

> How far from a TSS each peak is?  

<details><summary style="color: #ff7f00; font-weight: bold">Show code</summary><p>
```{r}
# Get Yeast TSS annotations from TxDb
yeast_genes <- GenomicFeatures::genes(TxDb.Scerevisiae.UCSC.sacCer3.sgdGene::TxDb.Scerevisiae.UCSC.sacCer3.sgdGene)
GenomeInfoDb::seqlevelsStyle(yeast_genes) <- GenomeInfoDb::seqlevelsStyle(ATAC_peaks)
yeast_TSSs <- GenomicRanges::resize(yeast_genes, 1, 'start')
# Convert Ensembl gene ID to human-readable gene names, using gprofiler2
yeast_IDs <- gprofiler2::gconvert(query = yeast_TSSs$gene_id, organism = 'scerevisiae')
head(yeast_IDs)
yeast_TSSs$gene_name <- yeast_IDs$name[match(yeast_TSSs$gene_id, yeast_IDs$input)]
# Annotate closest TSS for each ATAC peak
ATAC_peaks$associated_gene <- yeast_TSSs$gene_name[nearest(ATAC_peaks, yeast_TSSs, ignore.strand = TRUE)]
# Calculate distance to the nearest TSS for each ATAC peak
dists <- IRanges::distanceToNearest(ATAC_peaks, yeast_TSSs, ignore.strand = TRUE)
summary(mcols(dists)$distance)
hist(log10(mcols(dists)$distance), breaks = 300)
```
</p></details><br>

> Compare location of ATAC-seq peaks with that of ChIP-seq peaks

<details><summary style="color: #ff7f00; font-weight: bold">Show code</summary><p>
```{r}
ChIP_peaks <- GenomicRanges::reduce(do.call(c, lapply(
    list.files(pattern = '.*narrowPeak$', path = 'Share/day03/results/bwa/mergedLibrary/macs/narrowPeak', full.names = TRUE), rtracklayer::import
)))
dists <- IRanges::distanceToNearest(ATAC_peaks, ChIP_peaks)
summary(mcols(dists)$distance)
hist(log10(mcols(dists)$distance), breaks = 300)
```
</p></details><br>

## 3. Count ATAC fragments over ATAC peaks

In R, we can leverage Bioconductor classes and their optimized access to `.bam` files to efficiently count 
fragment coverage of ATAC-seq datasets over all the annotated ATAC-seq peaks. 

```{r, echo = FALSE, results = FALSE}
ATAC_counts <- readRDS('Share/day04/ATAC_counts.rds')
```
<details><summary style="color: #ff7f00; font-weight: bold">Show code</summary><p>
```{r}
# Define sample sheet
sampleTable <- data.frame(
    sample = gsub('_SRR.*', '', list.files(pattern = '_noChrM_q10.bam$', path = 'Share/day04/_bam-files/')), 
    timepoint = gsub('SacCer3_|_rep.*', '', list.files(pattern = '_noChrM_q10.bam$', path = 'Share/day04/_bam-files/')), 
    replicate = rep(c(1, 2), 5), 
    bam = list.files(pattern = '_noChrM_q10.bam$', path = 'Share/day04/_bam-files/', full.names = TRUE)
)
```
```{r, eval = FALSE}
# Import fragment counts over peaks 
ATAC_counts <- GenomicAlignments::summarizeOverlaps(, 
    features = ATAC_peaks, 
    reads = Rsamtools::BamFileList(sampleTable$bam), 
    ignore.strand = TRUE, 
    singleEnd = FALSE
)
SummarizedExperiment::colData(ATAC_counts) <- DataFrame(sampleTable)
```
```{r}
ATAC_counts
length(ATAC_counts)
dim(ATAC_counts)
dplyr::glimpse(ATAC_counts)
SummarizedExperiment::rowRanges(ATAC_counts)
SummarizedExperiment::colData(ATAC_counts)
```
</p></details><br>

## 4. Perform DA analysis

```{r, echo = FALSE, results = FALSE}
ATAC_peaks <- readRDS('Share/day04/ATAC_peaks.rds')
```

DESeq2 is probably the most widely used package to compute differential coverage (expression) of chromatin loci 
(genes). Using raw unnormalized counts, it performs the following steps: 

* It estimates the size factors for each library 
* It estimates the count-dependant dispersion 
* It fits a negative binomial generalized linear model to internally normalized counts
* Finally, it tests for differential coverage of genomic features (e.g. genes or peaks) and estimates the log2 fold-change between different conditions. 

<details><summary style="color: #ff7f00; font-weight: bold">Show code</summary><p>
```{r, eval = FALSE}
library(DESeq2)
# Perform DE analysis
dds <- DESeqDataSet(ATAC_counts, design = ~ timepoint) 
dds <- DESeq(dds)
# Prepare contrasts for timepoint pairwise comparison
timepoints <- c('00', '15', '30', '45', '60')
contrasts <- apply(combn(5, 2), 2, function(idx) {
    c('timepoint', timepoints[idx[2]], timepoints[idx[1]])
})
colnames(contrasts) <- apply(contrasts, 2, function(col) paste0(col[2], '_v_', col[3]))
contrasts
# Extract differential accessibility results
out.l2fc <- data.frame(matrix(nrow = nrow(dds), ncol = ncol(contrasts)))
colnames(out.l2fc) <- paste0('L2FC_', colnames(contrasts))
out.padj <- data.frame(matrix(nrow = nrow(dds), ncol = ncol(contrasts)))
colnames(out.padj) <- paste0('padj_', colnames(contrasts))
for (i in 1:ncol(contrasts)) {
    ddr <- results(dds, contrast = contrasts[,i])
    metadata(ddr)$alpha <- 0.01
    out <- as.data.frame(ddr)
    out.l2fc[,i] <- out[, 'log2FoldChange']
    out.padj[,i] <- out[, 'padj']
}
# Get regularized log-normalized accessibility counts
rlogs <- rlog(dds, blind = FALSE)
rlogs
rlogs <- assay(rlogs)
glimpse(rlogs)
colnames(rlogs) <- paste0(sampleTable$timepoint, "_rep", sampleTable$replicate)
rlogs <- matrix(unlist(lapply( 
    unique(sampleTable$timepoint), 
    function(x) { 
        x <- rowMeans(matrix(rlogs[,grep(x, colnames(rlogs))], nrow = nrow(rlogs))) 
    }
)), nrow = nrow(rlogs))
rownames(rlogs) <- paste0('peak_', 1:nrow(rlogs))
colnames(rlogs) <- paste0('rlog_', unique(sampleTable$timepoint))
# Collate all results together
results <- cbind(rlogs, out.l2fc, out.padj)
mcols(ATAC_peaks) <- cbind(mcols(ATAC_peaks), results)
```
```{r}
ATAC_peaks
glimpse(ATAC_peaks)
glimpse(ATAC_peaks@elementMetadata)
```
</p></details><br>

## 5. Check peaks DA between 2 timepoints

<details><summary style="color: #ff7f00; font-weight: bold">Show code</summary><p>
```{r}
# Get peaks whose accessibility increases between 0 and 15 min
g <- ATAC_peaks[which(ATAC_peaks$L2FC_15_v_00 > log2(1.5) & ATAC_peaks$padj_15_v_00 < 0.01)]
g[, 1:2]
# Get peaks whose accessibility increases between 0 and 60 min
g <- ATAC_peaks[which(ATAC_peaks$L2FC_60_v_00 > log2(1.5) & ATAC_peaks$padj_60_v_00 < 0.01)]
g[, 1:2]
```
</p></details><br>

## 6. Save analysis output

<details><summary style="color: #ff7f00; font-weight: bold">Show code</summary><p>
```{r}
# Export as RDS
saveRDS(ATAC_peaks, 'ATAC_peaks.rds')
# Export as GFF
attrs <- paste0(
    "ID=", ATAC_peaks$peakID,
    ";associated-gene=", ATAC_peaks$associated_gene, 
    ";rlog_00=", ATAC_peaks$rlog_00,
    ";rlog_15=", ATAC_peaks$rlog_15,
    ";rlog_30=", ATAC_peaks$rlog_30,
    ";rlog_45=", ATAC_peaks$rlog_45,
    ";rlog_60=", ATAC_peaks$rlog_60
)
GFF <- data.frame(
    seqname = as.character(seqnames(ATAC_peaks)),
    source = 'ATACseq_osmotic-stress-response', 
    feature = 'peak', 
    start = as.character(start(ATAC_peaks) + 1),
    end = as.character(end(ATAC_peaks)),
    score = '.', 
    strand = '*', 
    frame = '.',
    attribute = attrs
)
writeLines("##gffTags=on\n##displayName=Name\n##gff-version 3", 'ATAC_peaks.gff')
write.table(GFF, file = 'ATAC_peaks.gff', row.names = FALSE, col.names = FALSE, quote = FALSE, sep = "\t", append = TRUE)
```
</p></details><br>
